### 2015-03-08 doc/thesis/system.tex
Resizing margins and removing several typing rules from the text

Our core limits control flow to if and while statements;
it has explicit type annotations, explicit scope for variables,
explicit method declarations, and explicit method calls.
Here is a list of features that are not present in our reduced core:
\begin{itemize}
\item labels and goto statements are not present because they
are difficult to handle along with our simplified form of
\emph{flow typing}, and they are out of the scope for now;
\item explicit blocks are not present because we are already using
explicit scope for variables;
\item other loop structures such as repeat-until, numeric for,
and generic for are not present because we can use while to
express them;
\item fields in the table constructor are reduced to a single
form because we can use this form to express the missing forms; 
\item arithmetic operators are reduced to addition and unary minus
because the other operators have similar typing rules;
\item relational operators are reduced to equality and less than
because the other operators have similar typing rules;
\item bitwise operators are not present because they are similar
to integer arithmetic operators.
\end{itemize}

\begin{figure}[!ht]
\textbf{Abstract Syntax}\\
\dstart
$$
\begin{array}{rl}
s ::= & \;\; \mathbf{skip} \; | \;
s_{1} \; ; \; s_{2} \; | \;
\vec{l} = el \; | \;
m \; | \;
\mathbf{while} \; e \; \mathbf{do} \; s \; | \;
\mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
& | \; \mathbf{local} \; \vec{n{:}T} = el \; \mathbf{in} \; s \; | \;
\mathbf{local} \; \vec{n} = el \; \mathbf{in} \; s \; | \;
\mathbf{rec} \; n{:}T = f \; \mathbf{in} \; s\\
& | \; \mathbf{return} \; el \; | \;
\lfloor a \rfloor_{0}\\
e ::= & \;\; \mathbf{nil} \; | \;
k \; | \;
\lfloor {...} \rfloor_{1} \; | \;
n \; | \;
e_{1}[e_{2}] \; | \;
{<}T{>} \; n \; | \;
f \; | \;
\{ \; \vec{p} \; \} \; | \;
\{ \; \vec{p},me \; \}\\
& | \; e_{1} + e_{2} \; | \;
e_{1} \; {..} \; e_{2} \; | \;
e_{1} == e_{2} \; | \;
e_{1} < e_{2} \; | \;
e_{1} \; \mathbf{and} \; e_{2} \; | \;
e_{1} \; \mathbf{or} \; e_{2} \\
& | \; \mathbf{not} \; e \; | \;
- e \; | \;
\# \; e \; | \;
\lfloor a \rfloor_{1}\\
l ::= & \;\; n_{l} \; | \;
e_{1}[e_{2}] \; | \;
n[k] \; {<}T{>}\\
k ::= & \;\; \mathbf{false} \; | \;
\mathbf{true} \; | \;
{\it int} \; | \;
{\it float} \; | \;
{\it string}\\
p ::= & \;\; [e_{1}] = e_{2}\\
el ::= & \;\; \vec{e} \; | \;
\vec{e}, me\\
me ::= & \;\; a \; | \;
{...}\\
a ::= & \;\; e(el) \; | \;
e{:}n(el)\\
m ::= & \;\; \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}T}){:}S \; fb \; | \;
\mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}T},{...}{:}T){:}S \; fb\\
f ::= & \;\; \mathbf{fun} \; (\vec{n{:}T}){:}S \; fb \; | \;
\mathbf{fun} \; (\vec{n{:}t},{...}{:}T){:}S \; fb\\
fb ::= & \;\; s \;;\; \mathbf{return} \; el\\
n ::= & \;\; {\it name}\\
\end{array}
$$
\dend
\caption{Typed Lua abstract syntax}
\label{fig:syntax}
\end{figure}

Figure \ref{fig:syntax} presents the abstract syntax of core Typed Lua.
It splits the syntactic categories as follows:
$s$ are statements, $e$ are expressions, $l$ are left-hand values,
$el$ are expression lists, $k$ are literal constants, $p$ are table fields,
$me$ are expressions with multiple results, $f$ are function declarations,
$m$ are method declarations, $a$ are function and method applications,
$n$ are variable names, $T$ are first-level types, and $S$ are second-level types.
The notation $\vec{n{:}T}$ denotes the list $n_{1}{:}T_{1}, ..., n_{n}{:}T_{n}$,
where $n$ can be zero.

Our reduced core includes two statements for declaring local variables.
One with explicit type annotations, and another without type annotations.
While we use the former to show how our type system handles the declaration
of annoatted variables, we use the latter to show how our type system
handles the declaration of unannotated variables through local type inference
and also the introduction of projection types.

Our reduced core also includes a truncation operator $\lfloor \rfloor$ for
function applications, method applications, and the vararg expression.
We use $\lfloor a \rfloor_{0}$ to denote function and method applications
that produce no value because they appear as statements.
We use $\lfloor a \rfloor_{1}$ to denote function and method applications
that produce only one value, even if the function or method application
returns mutiple values.
We use $\lfloor {...} \rfloor_{1}$ to denote a vararg expression that
produces only one value, as it can appear in the middle of an
expression list.

We also include two kinds of type coercions in our core language:
the left-hand value $n[k] \; {<}T{>}$ and the expression ${<}T{>} \;n$.
Both allow the refinement of table types.
We also split variable names into two categories to have safe aliasing
of tables in the presence of refinement.
We use $n$ when they appear as expressions and $n_{l}$ when they
appear as left-hand values.

We present the typing rules as a deduction system for the typing relations
$\env_{1}, \penv \vdash s, \env_{2}$ and $\env_{1}, \penv \vdash e : T, \env_{2}$.
We use the first one for typing statements and the second one for typing expressions.
The first relation means that given a type environment $\env_{1}$
and a projection environment $\penv$, typing a statement $s$ produces
a new type environment $\env_{2}$.
The second relation means that given a type environment $\env_{1}$
and a projection environment $\penv$, typing an expression $e$ has
type $T$ and produces a new type environment $\env_{2}$.
We need two typing relations because only expressions have types,
though expressions and statements can modify a type environment.

Even though we can assign only first-level types to variables,
our type system should be able to project unions of first-level types
from unions of second-level types.
Thus, we need two environments because $\env$ maps variables to first-level types,
while $\penv$ maps variables to second-level types.
We will show how our type system uses the environment $\penv$ for
handling projection types.
We use $\env_{1}[n \mapsto T]$ to extend the environment $\env_{1}$
with the variable $n$ that maps to type $T$.
We use $\penv[x \mapsto S]$ to extend the environment $\penv$ with
the variable $x$ that maps to the second-level type $S$.

@@

Lua has multiple assignment, and the rule \textsc{T-ASSIGNMENT} uses
tuple types along with consistent-subtyping to type check this Lua feature:
\[
\begin{array}{c}
\mylabel{T-ASSIGNMENT}\\
\begin{array}{c}
\dfrac{\env_{1}, \penv \vdash el:S_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash \vec{l}:S_{2}, \env_{3} \;\;\;
       S_{1} \lesssim S_{2}}
      {\env_{1}, \penv \vdash \vec{l} = el,\env_{3}}
\end{array}
\end{array}
\]

As an example,
\[
x, y = 1, ``foo"
\]
type checks because
$1 \times ``foo" \times \Nil{*} \lesssim \Integer \times \String \times \Value{*}$,
assuming that $x$ and $y$ are in the environment with
types $\Integer$ and $\String$, respectively.

Our type system uses the type $\Value{*}$ to drop extra values
and the type $\Nil{*}$ to replace missing values.
The rule that type checks left-hand values list introduces the type
$\Value{*}$ to let the right-hand side produce more values than
expected in the left-hand side.
The rules that type check expression list introduce the type
$\Nil{*}$ to let the right-hand side produce fewer values than
expected in the left-hand side.
Our type system includes $\Nil{*}$ in the type of an expression list only
if its type does not end in another variadic type $t{*}$.
In the previous example, the rules \textsc{T-LHSLIST} and \textsc{T-EXPLIST2}
are the rules that introduce these types.
We define the rule \textsc{T-LHSLIST} as follows:
\[
\begin{array}{c}
\mylabel{T-LHSLIST}\\
\dfrac{\env, \penv \vdash l_{i}:T_{i}, \env_{i} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
       n = |\;\vec{l}\;|}
      {\env, \penv \vdash \vec{l}:T_{1} \times ... \times T_{n} \times \Value{*}, \env_{m}}
\end{array}
\]

First, this rule uses the same environment $\env$ to type check each
left-hand side value $l_{i}$.
Then, it takes each type $T_{i}$ and builds the tuple type for
the assignment rule, placing $\Value{*}$ in the end.
Table refinement can change the type environment while typing each left-hand value.
Thus, we use a partial auxiliary function \emph{merge} to collect
all the changes in a new environment $\env_{m}$, if there are no conflicts.

The rule \textsc{T-EXPLIST2} is similar to \textsc{T-LHSLIST}:
\[
\begin{array}{c}
\mylabel{T-EXPLIST2}\\
\dfrac{\env, \penv \vdash e_{i}:T_{i}, \env_{i} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
       n = |\;\vec{e}\;|}
      {\env, \penv \vdash \vec{e}:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{m}}
\end{array}
\]

We use \textsc{T-EXPLIST2} only when the last expression in the list
can produce a single value.
Rules \textsc{T-EXPLIST3} and \textsc{T-EXPLIST4} handle the case where
an expression list is just an expression that may produce multiple values:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-EXPLIST3}\\
\dfrac{\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n} \times \Void, \env_{2}}
      {\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{T-EXPLIST4}\\
\dfrac{\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n}{*}, \env_{2}}
      {\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n}{*}, \env_{2}}
\end{array}
\end{array}
\]

As an example, these rules allow our type system to type check the following example:
\[
x, y, z = f()
\]

In this example, we are assuming that $x$, $y$, $z$, and $f$ are in
the environment with types $\Integer$, $\String$, $\String \cup \Nil$, and
either $\Void \rightarrow \Integer \times \String \times \Void$ or
$\Value{*} \rightarrow \Integer \times \String \times \Nil{*}$,
depending on whether the programmer is using strict function types.
Thus, the example type checks because
\[
\Integer \times \String \times \Nil{*} \lesssim \Integer \times \String \times \String \cup \Nil \times \Value{*}
\]

Conversely,
\[
x = f()
\]
type checks because
\[
\Integer \times \String \times \Nil{*} \lesssim \Integer \times \Value{*}
\]

@@

Our abstract syntax reduces the syntactic forms of the table constructor
into two forms: $\{\;\vec{p}\;\}$ and $\{\;\vec{p},me\;\}$.
The first uses a list of table fields $p_{1}, ..., p_{n}$,
where $p$ is $[e_{1}] = e_{2}$ and $n$ can be zero.
The second uses a list of table fields and an expression that can
produce multiple values.
We did not include in the abstract syntax a table constructor that uses
lists of expressions, because we can express this form using lists of
table fields.
For instance, we can use the table constructor
$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ to express $\{ ``x", ``y", ``z" \}$.

@@

The rule \textsc{T-CONSTRUCTOR2} type checks a table type with a non-empty
list of table fields:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR2}\\
\dfrac{\env_{i}, \penv \vdash p_{i}:(K_{i},V_{i}), \env_{i+1} \;\;\;
       n = |\;\vec{p}\;| \;\;\;
       T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique} \;\;\;
       wf(T)}
      {\env_{1}, \penv \vdash \{\;\vec{p}\;\}:T, \env_{n+1}}
\end{array}
\]

First, the rule \textsc{T-CONSTRUCTOR2} uses the auxiliary relation
$\env_{1}, \penv \vdash p : (K,V), \env_{2}$ to type check each table
field $p_{i}$.
This auxiliary relation means that given a type environment $\env_{1}$
and a projection environment $\penv$, typing a table field $p$
produces the pair $(K,V)$ and a new type environment $\env_{2}$. 
In the pair $(K,V)$, $K$ is the key type, while $V$ is the value type.

Then, the rule \textsc{T-CONSTRUCTOR2} uses each pair $(K_{i},V_{i})$
to build the table type that express the type of the given constructor, and
uses the predicate \emph{wf} to check whether this table type is well-formed.
Formally, a table type is well-formed if it obeys the following rule:
\[
\forall i \not\exists j \; i \not= j \wedge K_{i} \lesssim K_{j}
\]

@@

The definition of the auxiliary relation that type checks table
fields would be straightforward if our table types allowed any
first-level type in the type of the keys.
The rules \textsc{T-FIELD1}, \textsc{T-FIELD2}, and \textsc{T-FIELD3}
check if the type of the key is a subtype of $\Boolean$, $\Number$,
or $\String$, respectively:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-FIELD1}\\
\dfrac{\env_{1}, \penv \vdash e_{1}:T_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash e_{2}:T_{2}, \env_{3} \;\;\;
       T_{1} \subtype \Boolean}
      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (T_{1},close(T_{2})), \env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-FIELD2}\\
\dfrac{\env_{1}, \penv \vdash e_{1}:T_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash e_{2}:T_{2}, \env_{3} \;\;\;
       T_{1} \subtype \Number}
      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (T_{1},close(T_{2})), \env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-FIELD3}\\
\dfrac{\env_{1}, \penv \vdash e_{1}:T_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash e_{2}:T_{2}, \env_{3} \;\;\;
       T_{1} \subtype \String}
      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (T_{1},close(T_{2})), \env_{3}}
\end{array}
\end{array}
\]

These rules use the function \emph{close} in the type of the value
because our type system allows only \emph{closed} table types to
appear in the type of the values.
The function \emph{close} promotes all table types to \emph{closed}.
We made this restriction because our type system does not keep track
of aliases to table fields.
This means that allowing \emph{unique}, \emph{open}, and \emph{fixed}
table types to appear in the type of the values would allow the creation
of unsafe aliases.

The rule \textsc{T-FIELD4} is a fallback for the previous rules:
\[
\begin{array}{c}
\mylabel{T-FIELD4}\\
\dfrac{\env_{1}, \penv \vdash e_{1}:T_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash e_{2}:T_{2}, \env_{3}}
      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (\Value,close(T_{2})), \env_{3}}
\end{array}
\]

The rule \textsc{T-FIELD4} shows that our type system uses the type $\Value$
as the type of the key whenever a key has a type that is neither a
subtype of $\Boolean$, nor a subtype of $\Number$, nor a subtype of $\String$.

As an example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
has type $\{``x":1, ``y":\{``z":2\}_{closed}\}_{unique}$ through the rules
\textsc{T-FIELD3} and \textsc{T-CONSTRUCTOR2}.

Typed Lua handles arrays as hashes that map integers to some type $T$.
In Lua, programmers often use the vararg expression to initialize arrays.
The rules \textsc{T-CONSTRUCTOR3} and \textsc{T-CONSTRUCTOR4} define
the behavior of the vararg expression for this case:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-CONSTRUCTOR3}\\
\dfrac{\env({...}) = T}
      {\env, \penv \vdash \{{...}\}:\{\Integer{:}T \cup \Nil\}_{unique}, \env}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-CONSTRUCTOR4}\\
\dfrac{\begin{array}{c}
       \env_{i}, \penv \vdash p_{i}:(K_{i},V_{i}), \env_{i+1} \;\;\;
       n = |\;\vec{p}\;| \;\;\;
       \env_{i+1}({...}) = T_{v}\\
       T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}, \Integer{:}T_{v} \cup \Nil\}_{unique} \;\;\;
       wf(T)
       \end{array}}
      {\env_{1}, \penv \vdash \{\;\vec{p},\;{...}\;\}:T, \env_{n+1}}
\end{array}
\end{array}
\]

The rule \textsc{T-CONSTRUCTOR3} type checks the case where we use
only the vararg expression inside a table constructor to initialize
an array.
If we assume that $...$ is in the environment and has type $\String$,
the table constructor $\{{...}\}$ has type $\{\Integer:\String \cup \Nil\}_{unique}$
through the rule \textsc{T-CONSTRUCTOR3}.

The rule \textsc{T-CONSTRUCTOR4} type checks the case where we use
the table constructor as a record that includes an array part.
If we assume that $...$ is in the environment and has type $\String$,
the table constructor $\{[``x"] = ``foo", [``y"] = \mathbf{true}, {...}\}$
has type $\{``x":``foo", ``y":\mathbf{true}, \Integer:\String \cup \Nil\}_{unique}$
through the rule \textsc{T-CONSTRUCTOR4}.
As another example, the rule \textsc{T-CONSTRUCTOR4} does not type check
the table constructor $\{ [1] = ``foo", [10] = \mathbf{true}, {...} \}$,
because the keys $1$ and $10$ overlap the array part,
and the resulting table type is not well-formed.

@@

We use the refinement of table types to handle the declaration of
new global variables.
In Lua, the assignment \texttt{v = v + 1} translates to
\texttt{\string_ENV["v"] = \string_ENV["v"] + 1} when \texttt{v}
is not a local variable, where \texttt{\string_ENV} is a table
that stores the global environment.
For this reason, Typed Lua treats accesses to global variables as field accesses
to an \emph{open} table in the top-level scope.
For instance,
\[
\string_ENV[``x"] \; {<}\String{>} = ``foo" \;;\; \string_ENV[``y"] \; {<}\Integer{>} = 1
\]
uses field assignment to add fields $``x"$ and $``y"$ to $\string_ENV$.
In this example and in the next examples, we assume that
$\string_ENV$ is in the environment and has type $\{\}_{open}$.
Therefore, after these field assignments $\string_ENV$ has type
$\{``x":\String, ``y":\Integer\}_{open}$.

@@

Aliasing an \emph{unique} or an \emph{open} table type produces a
\emph{closed} table type, as defined by rule \textsc{T-IDREAD}:
\[
\begin{array}{c}
\mylabel{T-IDREAD}\\
\dfrac{\env(n) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
      {\env, \penv \vdash n:close(T_{2}), \env[n \mapsto open(T_{1})]}
\end{array}
\]

In Section \ref{sec:fap} we will explain that \textsc{T-IDREAD} uses
the auxiliary function \emph{read} because it may be accessing an
identifier that is bound to a filter or projection type.
As we mentioned in Section \ref{sec:unions}, our type system includes
a small set of \texttt{type} predicates that allow programmers to
discriminate union types.
In Section \ref{sec:fap} we will also explain that our type system
uses filter and projection types in the definition of these predicates
to handle the discrimination of unions of first-level and second-level
types.

>>>>

Our core limits control flow to if and while statements;
it has explicit type annotations, explicit scope for variables,
explicit method declarations, and explicit method calls.
Here is a list of features that are not present in our reduced core:
\begin{itemize}
\item labels and goto statements are not present because they
are difficult to handle along with our simplified form of
\emph{flow typing}, and they are out of scope for now;
\item explicit blocks are not present because we are already using
explicit scope for variables;
\item other loop structures such as repeat-until, numeric for,
and generic for are not present because we can use while to
express them;
\item fields in the table constructor are reduced to a single
form because we can use this form to express the missing forms; 
\item arithmetic operators are reduced to addition only
because the other operators have similar typing rules;
\item relational operators are reduced to equality and less than
because the other operators have similar typing rules;
\item bitwise operators are reduced to bitwise and only
because the other operators have similar typing rules.
\end{itemize}

\begin{figure}[!ht]
\textbf{Abstract Syntax}\\
\dstart
$$
\begin{array}{rl}
s ::= & \;\; \mathbf{skip} \; | \;
s_{1} \; ; \; s_{2} \; | \;
\overline{l} = el \; | \;
m \; | \;
\mathbf{while} \; e \; \mathbf{do} \; s \; | \;
\mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
& | \; \mathbf{local} \; \overline{id{:}T} = el \; \mathbf{in} \; s \; | \;
\mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s \; | \;
\mathbf{rec} \; id{:}T = f \; \mathbf{in} \; s\\
& | \; \mathbf{return} \; el \; | \;
\lfloor a \rfloor_{0}\\
e ::= & \;\; \mathbf{nil} \; | \;
k \; | \;
id \; | \;
e_{1}[e_{2}] \; | \;
{<}T{>} \; id \; | \;
f \; | \;
\{ \; \overline{p} \; \} \; | \;
\{ \; \overline{p},me \; \} \; | \;
e_{1} + e_{2} \; | \;
e_{1} \; {..} \; e_{2} \\
& | \; e_{1} == e_{2} \; | \;
e_{1} < e_{2} \; | \;
e_{1} \;\&\; e_{2} \; | \;
e_{1} \; \mathbf{and} \; e_{2} \; | \;
e_{1} \; \mathbf{or} \; e_{2} \; | \;
\mathbf{not} \; e \; | \;
\# \; e \; | \;
\lfloor me \rfloor_{1}\\
l ::= & \;\; id_{l} \; | \;
e_{1}[e_{2}] \; | \;
id[k] \; {<}T{>}\\
k ::= & \;\; \mathbf{false} \; | \;
\mathbf{true} \; | \;
{\it int} \; | \;
{\it float} \; | \;
{\it string}\\
p ::= & \;\; [e_{1}] = e_{2}\\
el ::= & \;\; \overline{e} \; | \;
\overline{e}, me\\
me ::= & \;\; a \; | \;
{...}\\
a ::= & \;\; e(el) \; | \;
e{:}n(el)\\
m ::= & \;\; \mathbf{fun} \; id_{1}{:}id_{2} \; (\overline{id{:}T}){:}S \; fb \; | \;
\mathbf{fun} \; id_{1}{:}id_{2} \; (\overline{id{:}T},{...}{:}T){:}S \; fb\\
f ::= & \;\; \mathbf{fun} \; (\overline{id{:}T}){:}S \; fb \; | \;
\mathbf{fun} \; (\overline{id{:}T},{...}{:}T){:}S \; fb\\
fb ::= & \;\; s \;;\; \mathbf{return} \; el\\
id ::= & \;\; {\it name}\\
\end{array}
$$
\dend
\caption{Typed Lua abstract syntax}
\label{fig:syntax}
\end{figure}

Figure \ref{fig:syntax} presents the abstract syntax of core Typed Lua.
It splits the syntactic categories as follows:
$s$ are statements, $e$ are expressions, $l$ are left-hand values,
$el$ are expression lists, $k$ are literal constants, $p$ are table fields,
$me$ are expressions with multiple results, $f$ are function declarations,
$m$ are method declarations, $a$ are function and method applications,
$id$ are variable names, $T$ are first-level types, and $S$ are second-level types.
The notation $\overline{id{:}T}$ denotes the list $id_{1}{:}T_{1}, ..., id_{n}{:}T_{n}$,
where $n$ can be zero.

Our reduced core includes two statements for declaring local variables,
one with and another without type annotations.
While we use the former to show how our type system handles the declaration
of annoatated variables, we use the latter to show how our type system
handles the declaration of unannotated variables through local type inference
and also the introduction of projection types.

Our reduced core also includes a truncation operator $\lfloor \rfloor$ for
function applications, method applications, and the vararg expression.
We use $\lfloor a \rfloor_{0}$ to denote function and method applications
that produce no value because they appear as statements.
We use $\lfloor me \rfloor_{1}$ to denote function applications,
method applications, and vararg expressions that produce only one value,
even if they return mutiple values.

We also include two kinds of type coercions in our core language:
the left-hand value $id[k] \; {<}T{>}$ and the expression ${<}T{>} \;id$.
Both allow the refinement of table types.
We also split variable names into two categories to have safe aliasing
of tables in the presence of refinement.
We use $id$ when variable names appear as expressions and $id_{l}$ when
variable names appear as left-hand values.

We present the typing rules as a deduction system for the typing relations
$\env_{1}, \penv \vdash s, \env_{2}$ and $\env_{1}, \penv \vdash e : T, \env_{2}$.
We use the first one for typing statements and the second one for typing expressions.
The first relation means that given a type environment $\env_{1}$
and a projection environment $\penv$, we can check a statement $s$
producing a new type environment $\env_{2}$.
The second relation means that given a type environment $\env_{1}$
and a projection environment $\penv$, we can check that an expression $e$ has
type $T$ and produces a new type environment $\env_{2}$.

Even though we can assign only first-level types to variables,
functions and methods can return unions of second-level types,
and our type system should be able to project these unions of
second-level types into unions of first-level types.
To do that, we use projection types and the environment $\penv$ that
maps projection variables to second-level types, while the
environment $\env$ maps variables to first-level types.
We will show how our type system uses the environment $\penv$ for
handling projection types in Section \ref{sec:fap}.
We use $\env_{1}[id \mapsto T]$ to extend the environment $\env_{1}$
with the variable $id$ that maps to type $T$.
We use $\penv[x \mapsto S]$ to extend the environment $\penv$ with
the projection variable $x$ that maps to the second-level type $S$.

@@

Lua has multiple assignment, and our type system uses three
different kinds of typing rules to type check this feature.
It uses typing rules that type check the different forms
of expression lists that can appear in the right-hand side,
a typing rule that type checks a list of left-hand values,
and a general rule that uses consistent-subtyping to check
whether the type of the right-hand side is consistent with
the type of the left-hand side.

As an example, lets assume that $x$ and $y$ are variables in the
environment with types $\Integer$ and $\String$, the assignment
\[
x, y = 1, ``foo"
\]
type checks through the following rules:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-ASSIGNMENT}\\
\dfrac{\env_{1}, \penv \vdash el:S_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash \overline{l}:S_{2}, \env_{3} \;\;\;
       S_{1} \lesssim S_{2}}
      {\env_{1}, \penv \vdash \overline{l} = el,\env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-EXPLIST1}\\
\dfrac{\env, \penv \vdash e_{i}:T_{i}, \env_{i} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
       n = |\;\overline{e}\;|}
      {\env, \penv \vdash \overline{e}:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{m}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-LHSLIST}\\
\dfrac{\env, \penv \vdash l_{i}:T_{i}, \env_{i} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
       n = |\;\overline{l}\;|}
      {\env, \penv \vdash \overline{l}:T_{1} \times ... \times T_{n} \times \Value{*}, \env_{m}}
\end{array}
\end{array}
\]

The rule \textsc{T-ASSIGNMENT} is the general rule,
\textsc{T-EXPLIST1} is the rule that type checks an expression list
where all expressions can only produce a single value, and
\textsc{T-LHSLIST} is the rule that type checks a list of left-hand values.

Note that our type system uses the type $\Value{*}$ to drop extra values
and the type $\Nil{*}$ to replace missing values.
The rule that type checks lists of left-hand values introduces the type
$\Value{*}$ to let the right-hand side produce more values than
expected in the left-hand side.
The rules that type check expression lists introduce the type
$\Nil{*}$ to let the right-hand side produce fewer values than
expected in the left-hand side.

In Section \ref{sec:refinement} we will show that table refinement can
change the type environment while typing an expression or a left-hand value.
Thus, the rules that type check lists of expressions and lists of
left-hand values use a partial auxiliary function \emph{merge} to collect
all environment changes in a new environment $\env_{m}$, if there are no conflicts.

The previous example type checks through these rules because
\[
1 \times ``foo" \times \Nil{*} \lesssim \Integer \times \String \times \Value{*}
\]

As another example, lets assume that $x$, $y$, and $z$ are variables in
the environment with types $\Integer$, $\String$, and $\String \cup \Nil$,
the assignment
\[
x, y, z = 1, ``foo"
\]
type checks because
\[
\Integer \times \String \times \Nil{*} \lesssim \Integer \times \String \times \String \cup \Nil \times \Value{*}
\]

Conversely, the assignment
\[
x = 1, ``foo"
\]
type checks because
\[
\Integer \times \String \times \Nil{*} \lesssim \Integer \times \Value{*}
\]

Our type system includes $\Nil{*}$ in the type of an expression list only
if its type does not end in another variadic type $T{*}$.
For instance, rules \textsc{T-EXPLIST2} and \textsc{T-EXPLIST3} handle
the case where an expression list is just an expression that may produce
multiple values.
The former rule includes the type $\Nil{*}$, but the latter rule does not:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-EXPLIST2}\\
\dfrac{\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n} \times \Void, \env_{2}}
      {\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{T-EXPLIST3}\\
\dfrac{\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n}{*}, \env_{2}}
      {\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n}{*}, \env_{2}}
\end{array}
\end{array}
\]

@@

Our abstract syntax reduces the syntactic forms of the table constructor
into two forms: $\{\;\overline{p}\;\}$ and $\{\;\overline{p},me\;\}$.
The first uses a list of table fields $p_{1}, ..., p_{n}$,
where $p$ is $[e_{1}] = e_{2}$ and $n$ can be zero.
The second uses a list of table fields and an expression that can
produce multiple values.
We did not include in the abstract syntax a table constructor that uses
lists of expressions, because we can express this form using lists of
table fields.
For instance, we can use the table constructor
$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ to express $\{ ``x", ``y", ``z" \}$.

@@

The rule \textsc{T-CONSTRUCTOR2} type checks a table constructor with a non-empty
list of table fields:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR2}\\
\dfrac{\env_{i}, \penv \vdash p_{i}:(K_{i},V_{i}), \env_{i+1} \;\;\;
       n = |\;\vec{p}\;| \;\;\;
       T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique} \;\;\;
       wf(T)}
      {\env_{1}, \penv \vdash \{\;\vec{p}\;\}:T, \env_{n+1}}
\end{array}
\]

First, the rule \textsc{T-CONSTRUCTOR2} uses the auxiliary relation
$\env_{1}, \penv \vdash p : (K,V), \env_{2}$ to type check each table
field $p_{i}$.
This auxiliary relation means that given a type environment $\env_{1}$
and a projection environment $\penv$, we can check a table field $p$
producing a pair $(K,V)$ and a new type environment $\env_{2}$. 
In the pair $(K,V)$, $K$ is the key type, while $V$ is the value type.

Then, the rule \textsc{T-CONSTRUCTOR2} uses each pair $(K_{i},V_{i})$
to build the table type that express the type of a given constructor, and
uses the predicate \emph{wf} to check whether this table type is well-formed.
Formally, a table type is well-formed if it obeys the following rule:
\[
\forall i \not\exists j \; i \not= j \wedge K_{i} \lesssim K_{j}
\]

@@

Well-formed table types also do not allow \emph{unique} and
\emph{open} table types to appear in the type of the values.
We made this restriction because our type system does not keep
track of aliases to table fields.
This means that allowing \emph{unique} and \emph{open} table
types to appear in the type of the value would allow the
creation of unsafe aliases.

@@

We use the refinement of table types to handle the declaration of
new global variables.
In Lua, the assignment \texttt{v = v + 1} translates to
\texttt{\string_ENV["v"] = \string_ENV["v"] + 1} when \texttt{v}
is not a local variable, where \texttt{\string_ENV} is a table
that stores the global environment.
For this reason, Typed Lua treats accesses to global variables as field accesses
to an \emph{open} table in the top-level scope.
In the following examples we assume that $\string_ENV$ is in the
environment and has type $\{\}_{open}$. For instance,
\[
\string_ENV[``x"] \; {<}\String{>} = ``foo" \;;\; \string_ENV[``y"] \; {<}\Integer{>} = 1
\]
uses field assignment to add fields $``x"$ and $``y"$ to $\string_ENV$.
Therefore, after these field assignments $\string_ENV$ has type
$\{``x":\String, ``y":\Integer\}_{open}$.

@@

Aliasing an \emph{unique} or an \emph{open} table type can produce
either a \emph{closed} or a \emph{fixed} table type, depending on
the context that we are using a variable.
Rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2} define this behavior:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-IDREAD1}\\
\dfrac{\env(n) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
      {\env, \penv \vdash n:close(T_{2}), \env[n \mapsto open(T_{1})]}
\end{array}
\;
\begin{array}{c}
\mylabel{T-IDREAD2}\\
\dfrac{\env(n) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
      {\env, \penv \vdash n:fix(T_{2}), \env[n \mapsto fix(T_{1})]}
\end{array}
\end{array}
\]

Both rules use the auxiliary function \emph{read} because they may be
accessing an identifier that is bound to a filter or projection type.
As we mentioned in Section \ref{sec:unions}, our type system includes
a small set of \texttt{type} predicates that allow programmers to
discriminate union types, and our type system uses filter and projection
types in the definition of these predicates to handle the discrimination
of unions of first-level and second-level types.
We can define \emph{read} as follows:
\begin{align*}
read(\penv, \phi(T_{1},T_{2})) & = T_{2}\\
read(\penv, \pi_{i}^{x}) & = proj(\penv(x), i)\\
read(\penv, T) & = T
\end{align*}
